miniKanren with fair search strategies

### 2019-05-20 Paper/paper.tex
more fixes

Fig.~\ref{DFSi} shows parts that are later compared with other search 
strategies. We follow some conventions to name variables: variables bound to 
states are named \texttt{s}; variables bound to goals are named `g' (possibly 
with subscript); variables bound to search spaces have names ending with 
$^\infty$. The first function, \texttt{disj2}, implements binary disjunction. 
\texttt{append-inf} is its helper, which composes two disjunctive search 
spaces. The following function, \texttt{conj2}, implements binary conjunction. 
It applies the \emph{first} goal to the input state, then applies the second 
goal to states in the resulting search space. The latter process is done with a 
helper function. \texttt{append-map-inf} applies its input goal to states in 
its input search spaces and compose the resulting search spaces. It reuses 
\texttt{append-inf} for search space composition. The following two definitions 
say disjunctive operator and conjunctive operator are right-associative.

@@

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \disj{} and unfair 
\conj{}. The implementation of DFS$_\textrm{bi}$ differs from 
DFS$_\textrm{i}$'s 
in the \disj{} macro. We list the new \disj{} with its helper in 
Fig.~\ref{balanced-disj}. The helper \texttt{disj+} builds a balanced binary 
tree 
whose leaves are the goals and whose nodes are \texttt{disj2}s, hence the name 
of 
this search strategy. The first argument 
to \texttt{disj+} is a list of all goals. And the next two arguments accumulate 
goals in 
the left and right subtrees. The first clause says that when there is one goal, 
the tree is the goal itself. When there are more goals, the first argument is 
partitioned into two sub-lists. The partition is done by repetitively 
dispatching the first two goals (the last clause), until no goal remains (the 
second clause) or one goal remains (the third clause). In contrast, the \disj{} 
in DFS$_\textrm{i}$ constructs the binary tree with the same collection of leaf 
nodes but in a particularly unbalanced form.

@@

Fair DFS (\DFSf) has fair \disj{} and unfair \conj{}. The 
implementation of \DFSf{} differs from \DFSi{}'s in 
\texttt{disj2} (Fig.~\ref{fDFS}). \texttt{disj2} is changed to call a new and 
fair version of \texttt{append-inf}. \texttt{append-inf/fair} immediately calls 
its helper, \texttt{loop}, with the first argument, \texttt{s?}, set to 
\texttt{\#{}t}, which indicates that 
\texttt{s-inf} and \texttt{t-inf} haven't been swapped. The swapping happens at 
the third \texttt{cond} clause in the helper, where \texttt{s?} is updated 
accordingly. The first two \texttt{cond} clauses essentially copy the 
\texttt{car}s and stop recursion when one of the input spaces is obviously 
finite. The third clause, as we mentioned above, is just for swapping. When the 
fourth and last clause runs, we know that both \texttt{s-inf} and \texttt{t-inf} 
are ended with a thunk. In this case, a new thunk is constructed. The new thunk 
calls the driver recursively. Here changing the order of \texttt{t-inf} and 
\texttt{s-inf} won't hurt the fairness (though it will change the order of 
answers). We swap them back so that answers are produced in a more natural 
order.

@@

BFS has both fair \disj{} and fair \conj{}. Our implementation is based on 
DFS$_\textrm{f}$ (not DFS$_\textrm{i}$). To implement \BFS{} based on \DFSf{}, 
we need \texttt{append-map-inf/fair} in addition to \texttt{append-inf/fair}. 
The only difference between \texttt{append-map-inf/fair} and 
\texttt{append-map-inf} is that the former calls \texttt{append-inf/fair} 
instead of \texttt{append-inf}.

The implementation can be improved in two ways. First, as mentioned in 
section 2.2, BFS puts answers in bags and answers of the same cost are in the 
same bag. In this implementation, however, it is unclear where this information 
is recorded. Second, \texttt{append-inf/fair} is extravagant in memory usage. It 
makes $O(n+m)$ new \texttt{cons} cells every time, where $n$ and $m$ are the 
``length''s of input search spaces. We address these issues in the first 
subsection.

@@

As mentioned in section 2.2, BFS puts answers in bags and answers of the 
same cost are in the same bag. The cost
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are in the same bag), and the costs of answers in thunk are 
computed recursively then increased by one. It is even more subtle that
\texttt{append-inf/fair} and the \texttt{append-map-inf/fair} respects the cost 
information. We make these facts more obvious by changing the type of search 
space, modifying related function definitions, and introducing a few more 
functions.

@@

Functions related to the pure subset are listed in Fig.~\ref{BFS-opt} (the 
others in Fig.~\ref{BFS-opt-cont}). They are compared with 
\citeauthor{seres1999algebra}'s implementation later. The first three functions 
in Fig.~\ref{BFS-opt} are search space constructors. \texttt{none} makes an 
empty search space; \texttt{unit} makes a space from one answer; and 
\texttt{step} makes a space from a thunk. The remaining functions do the same 
thing as before.

Luckily, the change in \texttt{append-inf/fair} also fixes the miserable space 
extravagance -- the use of \texttt{append} helps us to reuse the first bag of 
\texttt{t-inf}.

\citet{kiselyov2005backtracking} has shown that a \emph{MonadPlus} hides in 
implementations of logic programming system. Our BFS implementation is not an 
exception: \texttt{append-map-inf/fair} is like \texttt{bind}, 
but takes arguments in reversed order; \texttt{none}, \texttt{unit}, and 
\texttt{append-inf/fair} correspond to \texttt{mzero}, \texttt{unit}, 
and \texttt{mplus} respectively.

@@

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s-inf} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s-inf} contains some 
answers, \texttt{ks} is called with the first answer and the rest space. 
Otherwise, \texttt{kf} is called with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is like an eliminator of 
search space, hence the name. The remaining functions do the same thing as 
before.

@@

In this section, we compare the pure subset of our optimized BFS with the BFS 
found in \citet{seres1999algebra}. We focus on the pure subset because 
Silvija's system is pure. Their system represents search spaces with streams of 
lists of answers, where each list is a bag.

To compare efficiency, we translate her Haskell code into Racket (See 
supplements for the translated code). The translation is direct 
due to the similarity in both logic programming systems and search space 
representations. The translated code is longer and slower than our BFS 
implementation. Details about 
difference in efficiency are in section 6.

@@

In this section, we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means running out of 
memory. The first two benchmarks are taken from 
\citet{friedman_reasoned_2018}. \texttt{reverso} is from 
\citet{rozplokhas2018improving}. Next two benchmarks 
about quine are modified from a similar test case in \citet{byrd2017unified}. 
The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absento}). Our version generates de 
Bruijnized expressions and prevent closures getting into list. The two 
benchmarks differ in the \conde{} clause order of their relational 
interpreters. 
The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is also inspired by \citet{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde{} clause order of their relational 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that DFS$_\textrm{i}$ would 
perform badly in the second case because elimination rules complicate the 
problem when running backward. The evaluation supports our conjecture.

In general, only DFS$_\textrm{i}$ and DFS$_\textrm{bi}$ constantly perform well. DFS$_\textrm{f}$ is just as 
efficient in all benchmarks but \texttt{very-recursiveo}. Both BFS have obvious 
overhead in many cases. Among the three variants of DFS (they all have unfair 
\conj{}), DFS$_\textrm{f}$ is most resistant to clause permutation, followd by DFS$_\textrm{bi}$ then 
DFS$_\textrm{i}$. Among the two implementation of BFS, ours constantly performs as well or 
better. Interestingly, every strategies with fair \disj{} suffers in 
\texttt{very-recursiveo} and DFS$_\textrm{f}$ performs well elsewhere. 
Therefore, this 
benchmark might be a special case. Fair \conj{} imposes overhead constantly 
except in \texttt{appendo}. The reason might be that strategies with fair 
\conj{} tend to keep more intermediate answers in the memory.

@@

Silvija et al \citet{seres1999algebra} also describe a breadth-first search 
strategy. We proof their BFS is equivalent to ours. But our code looks simpler 
and performs better in comparison with a straightforward translation of their 
Haskell code.

>>>>

Fig.~\ref{DFSi} shows parts that are later compared with other search 
strategies. We follow some conventions to name variables: variables bound to 
states are named \texttt{s}; variables bound to goals are named \texttt{g} 
(possibly with subscript); variables bound to search spaces have names ending 
with $^\infty$. The first function, \texttt{disj2}, implements binary 
disjunction. if and only if is its helper, which composes two disjunctive 
search spaces. The following function, \texttt{conj2}, implements binary 
conjunction. 
It applies the \emph{first} goal to the input state, then applies the second 
goal to states in the resulting search space. The latter process is done with a 
helper function. \texttt{append-map$^\infty$} applies its input goal to states 
in its input search spaces and compose the resulting search spaces. It reuses 
\texttt{append$^\infty$} for search space composition. The following and last 
two definitions say disjunctive operator and conjunctive operator are 
right-associative.

@@

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \disj{} and unfair 
\conj{}. The implementation of DFS$_\textrm{bi}$ differs from 
DFS$_\textrm{i}$'s in the \disj{} macro. We list the new \disj{} with its 
helper in Fig.~\ref{balanced-disj}. When there are one or more disjunctive 
goals, \disj{} builds a balanced binary tree whose leaves are the goals and 
whose nodes are \texttt{disj2}s, hence the name of this search strategy. 
The first ``argument'' to \texttt{disj+} is the goals. And the next two 
``arguments'' accumulate goals to be put in the left and right subtrees. When 
there is only one goal, the tree is the goal itself. This case is handled in 
the first clause. When there are more goals, we partition the list of goals 
into two sublists of roughly equal lengths and recur on the two sublists. The 
partition is done by repetitively dispatching the first two goals in the list 
to the accumulators, until only 
zero or one goal remains. The two base cases of the partition are handled by 
the second and the third clauses respectively. And the dispatching is done at 
the fourth and last clause. In contrast, the \disj{} in \DFSi{} constructs the 
binary tree in a particularly unbalanced form.

@@

Fair DFS (\DFSf) has fair \disj{} and unfair \conj{}. The 
implementation of \DFSf{} differs from \DFSi{}'s in 
\texttt{disj2} (Fig.~\ref{fDFS}). \texttt{disj2} is changed to call a new and 
fair version of \texttt{append$^\infty$}. \texttt{append$^\infty_{fair}$} 
immediately 
calls 
its helper, \texttt{loop}, with the first argument, \texttt{s?}, set to 
\texttt{\#{}t}, which indicates that 
\texttt{s$^\infty$} and \texttt{t$^\infty$} haven't been swapped. The swapping 
happens at 
the third \texttt{cond} clause in the helper, where \texttt{s?} is updated 
accordingly. The first two \texttt{cond} clauses essentially copy the 
\texttt{car}s and stop recursion when one of the input spaces is obviously 
finite. The third clause, as we mentioned above, is just for swapping. When the 
fourth and last clause runs, we know that both \texttt{s$^\infty$} and 
\texttt{t$^\infty$} are ended with a thunk and they have been swapped. In this 
case, a new thunk is constructed. Here changing the order of 
\texttt{t$^\infty$} and \texttt{s$^\infty$} won't hurt the fairness (though it 
will change the order of answers). We swap them back so that answers are 
produced in a more natural order.

@@

BFS has both fair \disj{} and fair \conj{}. Our implementation is based on 
DFS$_\textrm{f}$ (not DFS$_\textrm{i}$). To implement \BFS{} based on \DFSf{}, 
we need \texttt{append-map$^\infty_{fair}$} in addition to 
\texttt{append$^\infty_{fair}$}. 
The only difference between \texttt{append-map$^\infty$} and 
\texttt{append-map$^\infty_{fair}$} is that the latter calls 
\texttt{append$^\infty_{fair}$} instead of \texttt{append$^\infty$}.

The implementation can be improved in two ways. First, as mentioned in 
section 2.2, BFS puts answers in bags and answers of the same cost are in the 
same bag. In this implementation, however, it is unclear where this information 
is recorded. Second, \texttt{append$^\infty_{fair}$} is extravagant in memory 
usage. It 
makes $O(n+m)$ new \texttt{cons} cells every time, where $n$ and $m$ are the 
``length''s of input search spaces. We address these issues in the first 
subsection.

@@

As mentioned in section 2.2, BFS puts answers in bags and answers of the 
same cost are in the same bag. The cost
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are in the same bag), and the costs of answers in thunk are 
computed recursively then increased by one. It is even more subtle that
\texttt{append$^\infty_{fair}$} and the \texttt{append-map$^\infty_{fair}$} 
respects 
the cost 
information. We make these facts more obvious by changing the type of search 
space, modifying related function definitions, and introducing a few more 
functions.

@@

Functions related to the pure subset are listed in Fig.~\ref{BFS-opt} (the 
others in Fig.~\ref{BFS-opt-cont}). They are compared with 
\citeauthor{seres1999algebra}'s implementation in our proof. The first three 
functions in Fig.~\ref{BFS-opt} are search space constructors. \texttt{none} 
makes an empty search space; \texttt{unit} makes a space from one answer; and 
\texttt{step} makes a space from a thunk. The remaining functions do the same 
thing as before.

Luckily, the change in \texttt{append$^\infty_{fair}$} also fixes the miserable 
space 
extravagance -- the use of \texttt{append} helps us to reuse the first bag of 
\texttt{t$^\infty$}.

\citet{kiselyov2005backtracking} has shown that a \emph{MonadPlus} hides in 
implementations of logic programming system. Our BFS implementation is not an 
exception: \texttt{append-map$^\infty_{fair}$} is like \texttt{bind}, 
but takes arguments in reversed order; \texttt{none}, \texttt{unit}, and 
\texttt{append$^\infty_{fair}$} correspond to \texttt{mzero}, \texttt{unit}, 
and \texttt{mplus} respectively.

@@

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s$^\infty$} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s$^\infty$} contains 
some 
answers, \texttt{ks} is called with the first answer and the rest space. 
Otherwise, \texttt{kf} is called with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is like an eliminator of 
search space, hence the name. The remaining functions do the same thing as 
before.

@@

In this section, we compare the pure subset of our optimized BFS with the BFS 
found in \citet{seres1999algebra}. We focus on the pure subset because 
their system is pure. Their system represents search spaces with streams of 
lists of answers, where each list is a bag.

To compare efficiency, we translate their Haskell code into Racket (See 
supplements for the translated code). The translation is direct 
due to the similarity in both logic programming systems and search space 
representations. The translated code is longer and slower than our BFS 
implementation. Details about 
difference in efficiency are in section 6.

@@

In this section, we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means running out of 
memory. The first two benchmarks are taken from 
\citet{friedman_reasoned_2018}. \texttt{reverso} is from 
\citet{rozplokhas2018improving}. Next two benchmarks 
about quine are modified from a similar test case in \citet{byrd2017unified}. 
The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absent$^o$}). Our version generates de 
Bruijnized expressions and prevent closures getting into list. The two 
benchmarks differ in the \conde{} clause order of their relational 
interpreters. 
The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is also inspired by \citet{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde{} clause order of their relational 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that DFS$_\textrm{i}$ would 
perform badly in the second case because elimination rules complicate the 
problem when running backward. The evaluation supports our conjecture.

In general, only DFS$_\textrm{i}$ and DFS$_\textrm{bi}$ constantly perform well. DFS$_\textrm{f}$ is just as 
efficient in all benchmarks but \texttt{very-recursiveo}. Both BFS have obvious 
overhead in many cases. Among the three variants of DFS (they all have unfair 
\conj{}), DFS$_\textrm{f}$ is most resistant to clause permutation, followd by DFS$_\textrm{bi}$ then 
DFS$_\textrm{i}$. Among the two implementation of BFS, ours constantly performs as well or 
better. Interestingly, every strategies with fair \disj{} suffers in 
\texttt{very-recursiveo} and DFS$_\textrm{f}$ performs well elsewhere. 
Therefore, this 
benchmark might be a special case. Fair \conj{} imposes considerable overhead 
constantly except in \texttt{appendo}. The reason might be that strategies with 
fair \conj{} tend to keep more intermediate answers in the memory.

@@

Silvija et al \citet{seres1999algebra} also describe a breadth-first search 
strategy. We proof their BFS is equivalent to ours. But our code is shorter
and performs better in comparison with a straightforward translation of their 
Haskell code.
